Early Warning Signals and the Prosecutor's Fallacy

### 2012-04-19 fallacy.tex
minor
updates to fallacy

%\documentclass[authoryear,5p]{elsarticle}
\documentclass[authoryear,preprint,11pt]{elsarticle}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{lineno}
\linenumbers
\usepackage{subfigure}
\usepackage[pdftex]{color}
\definecolor{darkblue}{rgb}{0,0,0.5}
\definecolor{darkgreen}{rgb}{0,0.5,0}
%\usepackage[pdftex, colorlinks, citecolor=darkblue,linkcolor=darkgreen]{hyperref}
\usepackage[pdftex, colorlinks]{hyperref}
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
\newcommand{\ud}{\mathrm{d}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\C}{\mathrm{Cov}}
\newcommand{\V}{\mathrm{Var}}

\newcommand{\cb}[1]{{\it \color{darkgreen} (#1)}}

@@

 \section{Introduction}

@@

Though this system can be forced through a saddle-node bifurcation by
increasing values of either $e$ or $a$, all parameters are held fixed.
The simulation starts from the positive equilibrium population size.
Though the chance of a transition across the allee threshold in any 
given timestep are small, given enough time this system will eventually
experience such a rare event driving the population extinct.  We evolved
each replicate 50000 timesteps, sampling the system every 50 timesteps.  
In this time window 155 of the 1000 replicates experience population collapse.  
Nine of these replicates can be seen in Figure~\ref{fig:replicate_crashes},
as an illustration.

@@

To determine whether or not the variance or autocorrelation increased 
over time in these 155 replicates, we computed Kendall's $\tau$ for each of the
warning signal pattern~\citep{Dakos2008, Dakos2010}.  Positive values of $\tau$ 
indicate an increasing trend in the warning signal.  The distribution of $\tau$ 
values observed across these replicates is shown in Figure~\ref{fig:indicator}.
More positive values indicate stronger trends. Assessing the statistical significance
of a given value requires an appropriate null model, which can be difficult 
to do~\citep{Dakos2008}. As these examples have been selected from a system
that is not approaching a bifurcation and therefore should not exhibit any 
early warning sign, even the very strong values of $\tau$ should be considered 
as arising from the null model.

We compared this approach to a model-based approach

  \begin{figure}[ht]
    \begin{center}
      \includegraphics{replicate_crashes}
    \end{center}
    \caption{Nine of the 155 replicates that experience a critical transition by chance during the course of the simulation.}
    \label{fig:replicate_crashes}
  \end{figure}

  \begin{figure}[ht]
    \begin{center}
      \includegraphics{fallacy}
    \end{center}
    \caption{The distribution of three early warning indicators on replicates conditionally selected for having collapsed by chance in simulations.}
    \label{fig:indicator}
  \end{figure}

 \section{Discussion}

 \section{References}%bibliography
 \bibliography{boettiger}
 \end{document}

>>>>

%\documentclass[authoryear,5p]{elsarticle}
\documentclass[authoryear,review,12pt]{elsarticle}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{lineno}
\linenumbers
\usepackage{float}
\usepackage{subfigure}
\usepackage[pdftex]{color}
\definecolor{darkblue}{rgb}{0,0,0.5}
\definecolor{darkgreen}{rgb}{0,0.5,0}
\usepackage[pdftex, colorlinks]{hyperref}
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
\newcommand{\ud}{\mathrm{d}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\C}{\mathrm{Cov}}
\newcommand{\V}{\mathrm{Var}}
\newcommand{\cb}[1]{{\it \color{darkgreen} (#1)}}

%% Redefines the elsarticle footer
\makeatletter
\def\ps@pprintTitle{
\let\@oddhead\@empty
\let\@evenhead\@empty
\def\@oddfoot{\it \hfill\today}
\let\@evenfoot\@oddfoot}
\makeatother

@@

\section{Introduction}
In a criminal trial, the prosecution may introduce associative evidence such as 
blood testing to establish a match between the perpetrator at the scene of the
crime and a suspect on the defense stand.  Statistical testimony may accompany 
such evidence reflecting the incidence rate of the matching characteristic -- 
for instance, the frequency of individuals in the population with the matching
blood type.  Such arguments may be misleading when they fail to reflect the 
probability conditional on the manner in which the suspect was identified in 
the first place.  If the suspect has been selected partly on the basis of belonging
to demographic in which the characteristic is much more common than in the 
population as a whole, these odds can be misleading.  Failing to recognize 
these conditional probabilities leads to the ``Prosecutor's Fallacy'' -- the argument
that the suspect is the perpetrator on the basis of the statistical probability 
of a match without currectly identifying the baseline probability for such a match
\citep{Thompson1987}.

Statisical patterns associated with sudden, catastrophic transitions or ``tipping points''
could play an important role as early warning signs or leading indicators of
these shifts, which may give us precious time to react and avert them~\citep{Scheffer2009, Lenton2011}.
Historical examples of sudden transitions taken from the paleoclimate record
provide an important way to test and evaluate potential leading indicator methods,
and have been widely used for this purpose~\citep{Livina2007,Dakos2008,Lenton2012,Ditlevsen2010,Guttal2008,Thompson2010} 
Similarly, it has been suggested that data gathered on ecological systems such as lakes
that were monitored before the collapsed, or grasslands subjected to overgrazing,
could contain data that could help reveal when similar systems are approaching 
a tipping point~\citet{Inman2011}.  This is a silver lining to the clouds of past
catastrophes -- we may be able to learn from our mistakes.

In testing methods for early warning signals against historical examples of transitions,
we must be careful to avoid falling into the Prosecutor's Fallacy.  Sudden transitions
can occur for many reasons that do not involve the bifurcation proccesses on which
most early warning indicators are predicated~\citep{Hastings2010}, including purely
noise-induced transitions for which early warning signs will not exist~\citep{Ditlevsen2010, Lenton2011}.

% Explain variance, autocorrelation, tau.

\emph{Add more context \dots }

@@

Though this system can be forced through a saddle-node bifurcation by
increasing values of either $e$ or $a$, in these simulations all parameters
are held constant and no early warning signal is anticipated.
The simulation starts from the positive equilibrium population size.
Though the chance of a transition across the allee threshold in any 
given timestep are small, given enough time this system will eventually
experience such a rare event driving the population extinct.  We evolved
each replicate 50000 timesteps, sampling the system every 50 timesteps.  
In this time window 155 of the 1000 replicates experience population collapse.  
Nine of these replicates can be seen in Figure~\ref{fig:replicate_crashes},
as an illustration.

  \begin{figure}[H]
    \begin{center}
      \includegraphics{replicate_crashes.pdf}
    \end{center}
    \caption{Example trajectories for nine of the 155 replicates that 
             experience a critical transition by chance during the course
             of the simulation.}
    \label{fig:replicate_crashes}
  \end{figure}

@@

To determine whether or not the variance or autocorrelation increased 
over time in these 155 replicates, we computed Kendall's $\tau$ for each of the
warning signal pattern~\citep{Dakos2008, Dakos2010}.  Positive values of $\tau$ 
indicate an increasing trend in the warning signal.  The distribution of $\tau$ 
values observed across these replicates is shown in Figure~\ref{fig:indicator}.
More positive values indicate stronger trends. Assessing the statistical significance
of a given value requires an appropriate null model, which can be difficult 
to do~\citep{Dakos2008}. As these examples have been selected from a system
that is not approaching a bifurcation and therefore should not exhibit any 
early warning sign. %Even the very strong values of $\tau$ should be considered 
%as arising from the null model.

\begin{figure}[H]
  \begin{center}
    \includegraphics{indicators.pdf}
  \end{center}
  \caption{The distribution of two early warning indicators on replicates conditionally selected for having collapsed by chance in simulations.}
  \label{fig:indicator}
\end{figure}

For each of these replicates we also take a model-based approach, estimating 
parameters for an approximate linear model of the system approaching a
saddle node bifurcation.

\begin{equation}
  dX = \sqrt{ r_t } (\phi(r_t) - X_t) dt +
  \sigma\sqrt{\phi(r_t) } d B_t \label{LSN}
\end{equation}

where \( \phi(r_t) = \sqrt{r_t} +\theta \) is the equilbirium postition.  
This model emerges from the canonical form of the saddle-node bifurcation,

\begin{equation}
  \frac{dX}{dt} = r - x^2
\end{equation}

Where we introduce an explicit parameter $\theta$ to account for rescaling
to the units of the data,

\begin{equation}
   \frac{dX}{dt} = r - (x-\theta)^2
\end{equation}

and then consider the corresponding stochastic process where the noise is
proportional to the square root of the mean, as is the case for demographic
noise processes,

\begin{equation}
  \frac{dX}{dt} = r - (x-\theta)^2 + \sqrt{ \beta (r + (x-\theta)^2 } dB_t
\end{equation}

$r$ remains the bifurcation parameter which we imagine changing with time,
$r_t$, then linearizing the above model in neighboorhood of
\(\hat x = \sqrt{r_t} +\theta =: \phi(r_t)\), we recover Equation~\eqref{LSN}.
These models can be both be fit to the data by maximum likelihood estimation
based on the equation for the evolution of their moments, as
in~\citet{Boettiger2012b}.

Though this approximation is more general than the original Markov process,
it is also easier to estimate reliably from the data since it describes only 
the process near the equilibrium where the data is avialable.

We assume that $r_t$ is varying at constant rate, according to the model

\begin{equation}
 r_t = m t + r_0 
\end{equation}

and estimate $m$, $r_0$, $\theta$, and $\beta$ by maximum likelihood on each
of the simulated datasets.  Estimates of $m < 0 $ are expected in systems 
approaching a bifurcation, while for stable systems $m$ should be approximately zero.
Of the 155 simulations, only six had estimates with machine precision different
from zero, the greatest magnitude among these was $-7.1 \times 10^{-12}$,
which would still correspond to a negilible change in $r_0$ over the 50,000 timesteps
of the simulation.  Consequently, the model-based estimation shows no 
evidence of bias on data that has been selected conditional on collapse.

 \section{Discussion}
At first it seems tempting to argue that this bias is in fact valuable -- each
of these systems did indeed collapse, so the fact that they show increased 
probability to exhibit warning signals could be taken as a successful detection.
Unfortunately, this reasoning is flawed. At the moment the forecast is made,
these systems are no more likely to crash than their counterparts from the sample
that did not collapse, just as a fair coin is still a fair coin after a string of
consecutive ``heads.''  Taking a closer look at the patterns involved we can
begin to understand why common indicators such as autocorrelation and variance
can be misleading.

Systems that crash by chance excursion beyond the tipping point

Systems collapsing by chance do so from a string of unexpected events

Discuss the pattern of a system crashing by chance -- why this looks like increased autocorrelation and increased variance.

Discuss why this cannt be considered a successful detection.

These are not more likely to crash than another system.

Not a good test of the method's ability to detect bifurcations.

 \section{Acknowledgements}
This research was supported by funding from NSF Grant EF 0742674 to AH
and a Computational Sciences Graduate Fellowship from the Department of Energy grant DE-FG02-97ER25308 to CB. 
 \section{References}%bibliography
 \bibliography{fallacy}
 \end{document}
